---
title: "Final Project Analysis"
output: html_notebook
---

** Load packages
```{r}
library(mongolite) # for connecting with Mongo DB
library(e1071) # for Naive Bayes
library(hashmap) # for converting zipcodes to neighborhoods
library(class) # for knn
```

** Utility functions
```{r}
custMad <- function(pred, actual) {
  # returns the mad given the column of actual values and the column of the predicted values
  # args:
  # - Vector[double] pred; the predicted values from the model of the sale prices
  # - double givenM; the mean of actual values the predicted values are compared to
  # return double mad;
  mad <- sum(abs(pred - actual)) / length(pred)
  return(mad)
}
```

STEP 1: LOAD DATA

Load the proper collection stored in mongo
```{r}
mongo_conn <- mongo(collection = "apts", db = "apt_db")
test_apts <- mongo_conn$find('{}')
```

Load all of the one bedroom apartments into a dataframe
```{r}
one_bed_conn <- mongo(collection = "one_beds", db = "apt_db")
raw_one_beds <- one_bed_conn$find('{}')
```

Load all of the two bedroom apartments into a dataframe
```{r}
two_bed_conn <- mongo(collection = "two_beds", db = "apt_db")
raw_two_beds <- two_bed_conn$find('{}')
```

Load all of the three bedroom apartments into a dataframe
```{r}
three_bed_conn <- mongo(collection = "three_beds", db = "apt_db")
raw_three_beds <- three_bed_conn$find('{}')
```

Load all of the four and up bedroom apartments into a dataframe
```{r}
four_plus_conn <- mongo(collection = "four_up_beds", db = "apt_db")
raw_four_beds <- four_plus_conn$find('{}')
```

STEP 2: APPEND AND CLEAN DATA

Append dataframes for all bedrooms
```{r}
raw_apts <- do.call("rbind", list(raw_one_beds, raw_two_beds, raw_three_beds, raw_four_beds))
```

Clean apartment data
```{r}
apts <- raw_apts

# convert availability to just months or AvailableNow
apts$Available <- gsub('[[:digit:]]+', '', apts$Available)
apts$Available <- gsub("AvailableNow", "Now", apts$Available)

# alter column data types
apts <- apts[-which(apts$Address=="Boston"),]
apts$SqFt <- gsub("SqFt", "", apts$SqFt)
apts$SqFt <- as.numeric(gsub(",", "", apts$SqFt))
apts$Deposit <- as.numeric(gsub(",", "", apts$Deposit))
apts$Rent <- as.numeric(apts$Rent)
apts$Beds <- factor(apts$Beds)
apts$Baths <- factor(apts$Baths)
apts$Available <- factor(apts$Available)
apts$LeaseLength <- factor(apts$LeaseLength)
apts$City <- factor(apts$City)
apts$State <- factor(apts$State)
apts$Zip <- factor(apts$Zip)

# expand address
apts$Address <- gsub('([[:upper:]])', ' \\1', apts$Address)

# change row names and order
apts <- apts[,c("Address", "City", "State", "Zip", "Unit#", "Rent", "Beds", "Baths", "Available", "LeaseLength", "Deposit", "SqFt")]
names(apts) <- c("Address", "City", "State", "Zip", "Unit", "Rent", "Beds", "Baths", "Available", "LeaseLength", "Deposit", "SqFt")
```

Remove obviously fake information based on industry knowledge / guessing
This process is not done using formal outlier detection as it is possible that outliers are legitimate values
```{r}
# impute deposits that are less than $50 as NA
apts$Deposit[apts$Deposit < 50] <- NA

# impute SqFt where less than 50 or greater than 9950 as NA 
apts$SqFt[apts$SqFt < 50 | apts$SqFt > 9950] <- NA
```

Explore the data
```{r}
summary(apts)
```

STEP 3: PREDICT SQUARE FOOTAGE USING A MULTIPLE REGRESSION

View relationsips between square footage and other variables
```{r}
plot(apts$Rent, apts$SqFt)
abline(lm(SqFt ~ Rent, apts))
plot(apts$Beds, apts$SqFt)

# as expected, both variables show strong positive correlation with SqFt, thus should be included in any model
```

Separate all apartments into those with SqFt info and those without
```{r}
rowsNa <- which(is.na(apts$SqFt))
wSF <- apts[-rowsNa,]
woSF <- apts[rowsNa,]
```

Run a linear regression to model SqFt as a function of rent, beds, and baths
```{r}
# address, city, state and unit were not included because they're essentially uniquely identifying
# Used backwards elimination to remove Lease Length
sqft_model <- lm(SqFt ~ Rent + Beds + Baths + Zip, data = wSF)
summary(sqft_model)
```

Use the SqFt model to impute NAs in the SqFt column
```{r}
# create dataframe for apartments including imputed values and engineered fields
# dataframe does not inlude the 'Deposit' fields because most values are missing and its information is irrelevant regardless becuase it is almost always the same as rent
eng_apts <- apts[,!(names(apts) == "Deposit")]
eng_apts$SqFt[which(is.na(eng_apts$SqFt))] <- predict(sqft_model, newdata = eng_apts[which(is.na(eng_apts$SqFt)),], type = "response")

```

Create a derived feature 'Size' that describes an apartment as 'small', 'average', or 'large'
```{r}
# for each number of beds:
# find the lower and upper quartiles for sqft
# if sqft is below the 1st quartile then it is 'small'
# if sqft is between the 1st and 3rd quartile then it is 'average'
# if sqft is greater than the 3rd quartile then it is 'large'

sorted_beds <- sort(unique(eng_apts$Beds), decreasing = FALSE)
eng_apts$Size <- NA
for(bed in sorted_beds) {
  next_sqft <- eng_apts$SqFt[which(eng_apts$Beds==bed)]
  quart <- quantile(next_sqft, c(.25, .75), type = 1)
  quart <- unname(quart)
  eng_apts$Size[which(eng_apts$Beds==bed)] = sapply(eng_apts$SqFt[which(eng_apts$Beds==bed)], function(x) 
  {if(x < quart[1]) {
    return("Small")
    } else if(x >= quart[1] & x <= quart[2]) {
      return("Average")
    } else if (x > quart[2]){
      return("Large")
    } else {
      return(NA)
    }})
}
eng_apts$Size <- factor(eng_apts$Size)
```

Create a derived feature 'Price' that describes an apartment as 'cheap', 'average', or 'expensive'
```{r}
# for each number of beds:
# find the lower and upper quartiles for rent
# if rent is below the 1st quartile then it is 'cheap'
# if rent is between the 1st and 3rd quartile then it is 'average'
# if rent is greater than the 3rd quartile then it is 'expensive'

sorted_beds <- sort(unique(eng_apts$Beds), decreasing = FALSE)
eng_apts$Price <- NA
for(bed in sorted_beds) {
  next_price <- eng_apts$Rent[which(eng_apts$Beds==bed)]
  quart <- quantile(next_price, c(.25, .75), type = 1)
  quart <- unname(quart)
  eng_apts$Price[which(eng_apts$Beds==bed)] = sapply(eng_apts$Rent[which(eng_apts$Beds==bed)], function(x) 
  {if(x < quart[1]) {
    return("Cheap")
    } else if(x >= quart[1] & x <= quart[2]) {
      return("Average")
    } else if (x > quart[2]){
      return("Expensive")
    } else {
      return(NA)
    }})
}
eng_apts$Price <- factor(eng_apts$Price)
```

STEP 3: PREDICT NEIGHBORHOOD OF EACH APARTMENT

Generate a table of zip codes and their corresponding neighborhood taken from Boston public data
This will be used for training and validation of the model
```{r}
zips <- read.csv("/Users/cccdenhart/Documents/DS4100_Workspace/FinalProject/boston-office-by-zipcode.csv")[,3:4]
names(zips) <- c("Zip", "Neighborhood")

# clean the zipcodes
zips$Zip <- gsub(" ", "", zips$Zip)
zips$Zip <- gsub("\\*", "", zips$Zip)

# clean the neighborhoods
zips$Neighborhood <- gsub("Boston ", "", zips$Neighborhood)
zips$Neighborhood <- gsub("\\(", "", zips$Neighborhood)
zips$Neighborhood <- gsub("\\)", "", zips$Neighborhood)
zips$Neighborhood <- gsub("Roxbury Mission Hill", "Mission Hill", zips$Neighborhood)
zips$Neighborhood <- gsub("Dorchester Uphams Corner", "Dorchester", zips$Neighborhood)
zips$Neighborhood <- gsub("Dorchester Center Four Corners only", "Dorchester", zips$Neighborhood)
zips$Neighborhood <- gsub("Dorchester Grove Hall", "Dorchester", zips$Neighborhood)
zips$Neighborhood <- gsub("Dorchester Center excluding Four Corners", "Dorchester", zips$Neighborhood)
zips$Neighborhood <- gsub("Faneuil Hall marketplace/ north end", "North End", zips$Neighborhood)
zips$Neighborhood <- gsub("downtown crossing/ financial district", "Downtown", zips$Neighborhood)
zips$Neighborhood <- gsub("Hanover Street north end", "North End", zips$Neighborhood)
zips$Neighborhood <- gsub("Chinatown/ Tufts Medical", "Chinatown", zips$Neighborhood)
zips$Neighborhood <- gsub("Boston", "Seaport", zips$Neighborhood)
zips$Neighborhood <- gsub("South Seaport", "Seaport", zips$Neighborhood)
zips$Neighborhood <- gsub("East Seaport", "East Boston", zips$Neighborhood)

# clean the entire dataframe
zips <- zips[-c(which(zips$Zip==""), which(zips$Zip=="Thiszipcodeiscoveredbytwoareaoffices.")),]
extraZips <- data.frame(c("02118", "02145", "02119"), c("South End", "Somerville", "Roxbury"))
names(extraZips) <- c("Zip", "Neighborhood")
zips <- rbind(zips, extraZips)
zips
```

Add a neighborhood column to the data frame
```{r}
eng_apts$Neighborhood <- NA
find_neigh <- function(zip) {
  neigh <- zips$Neighborhood[which(zips$Zip==zip)][1]
  return(neigh)
}
eng_apts$Zip <- as.character(eng_apts$Zip)
eng_apts$Neighborhood <- sapply(eng_apts$Zip, find_neigh)
eng_apts$Neighborhood[which(eng_apts$Neighborhood=="character(0)")] <- NA
eng_apts$Neighborhood <- factor(eng_apts$Neighborhood)
names(eng_apts)
```

Get training and testing datasets from eng_apts to be used for a Naive-Bayes classifier
```{r}
nb_rows <- sample(x = nrow(eng_apts), size = nrow(eng_apts) * .6)
nb_train <- eng_apts[nb_rows,c("Price", "Beds", "Baths", "Available", "LeaseLength", "Size", "Neighborhood")]
nb_test <- eng_apts[-nb_rows,c("Price", "Beds", "Baths", "Available", "LeaseLength", "Size", "Neighborhood")]
```

Implement Naive-Bayes to predict an apartment's neighborhood using the training data
```{r}
nb_neighborhood <- naiveBayes(Neighborhood ~ ., data = nb_train)
```

Determine accuracy of the Naive-Bayes model using the testing data
```{r}
pred_test <- predict(nb_neighborhood, newdata = nb_test)
nb_correct <- length(which(pred_test == nb_test$Neighborhood))
nb_pct_corr <- nb_correct / nrow(nb_test)
nb_pct_corr
```

Get training and testing datasets from eng_apts to be used for a KNN model
```{r}
knn_rows <- sample(x = nrow(eng_apts), size = nrow(eng_apts) * .6)
knn_train <- eng_apts[knn_rows,c("Rent", "Beds", "Baths", "SqFt")]
train_classes <- eng_apts[knn_rows, "Neighborhood"]
knn_test <- eng_apts[-knn_rows,c("Rent", "Beds", "Baths", "SqFt")]
test_classes <- eng_apts[-knn_rows, c("Neighborhood")]
```

Implement KNN to predict an apartment's neighborhood using the training data
```{r}
init_k <- floor(sqrt(nrow(eng_apts)))
allK <- c((init_k-5):init_k,(init_k+1):(init_k+5))
accuracies <- c()
for(i in allK) {
  pred <- knn(knn_train, knn_test, train_classes, k=i)
  accuracy <- length(which(pred == test_classes)) / length(test_classes)
  accuracies <- c(accuracies, accuracy)
}

opt_k <- allK[which(accuracies==max(accuracies))]
knn_neighborhood <- knn(knn_train, knn_test, train_classes, k = opt_k)
```

Assess the accuracy of the KNN model using the testing data
```{r}
knn_correct <- length(which(knn_neighborhood==test_classes))
knn_pct_corr <- knn_correct / nrow(knn_test)
knn_pct_corr
```

Comparison of models
```{r}
# The two models produced very similarly poor results as they both were able to select only about 22% of correct neighborhoods.
# This is likely due to the fact that neighborhoods cannot be well identified simply by rent prices and size of apartments.
# For these models to be more acccurate, more data would have to be acquired about who posted the advertisement and who has responded to the advertisement
```

Predict rent price using a multiple regression
```{r}
rent_model <- lm(Rent ~ Zip + Beds + Baths + LeaseLength + SqFt, data = eng_apts)
summary(rent_model )
```

Evaluation of rent model fit
```{r}
rent_data <- eng_apts[,c("Zip", "Beds", "Baths", "LeaseLength", "SqFt")]
rent_data$LeaseLength[which(is.na(rent_data$LeaseLength))] <- 12
rent_real <- eng_apts[,"Rent"]
rent_pred <- predict(rent_model, rent_data)
mad <- custMad(rent_pred, rent_real)
mad
```


Export table as CSV
```{r}
write.csv(eng_apts, file = "/Users/cccdenhart/Documents/DS4100_Workspace/FinalProject/FinalTable.txt", row.names = F, col.names = F)
```

```{r}
test <- read.table("/Users/cccdenhart/Documents/DS4100_Workspace/FinalProject/FinalTable.txt", , sep=",", stringsAsFactors = F)
```

